RVD2: An ultra-sensitive variant detection model for heterogeneous population sequencing

### 2013-10-03 doc/bioinformatics2013/RVD2.tex
edits to manuscript.

\usepackage{verbatim}
\usepackage{amssymb}
\usepackage{amsmath}
%\usepackage{epstopdf}
%\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

@@

\section{Model structure}\

Given a data matrix $r \in \RR^{J \times N}$ where the element $r_{ji}$ is the count of the reads that contain a non-reference base at position $j$ in experimental replicate $i$. The total count of reads at position $i$ in replicate $i$ is $n_{ji}$. A typical data set is comprised of case and control data sets, $\{ r^{\text{case}} , n^{\text{case}} \}$ and  $\{ r^{\text{control}} , n^{\text{control}} \}$.

We have developed a hierarchical Bayesian model for this data set because of the nested structure of noise in the data. There is sampling error due to the fact that only a finite number of reads aligned and mapped for each location. There is inter-replicate errors due to experimental protocol and reproducibility. Finally there is sequence context error due to the fidelity of the DNA polymerase enzymes in synthesizing a base. These errors have been shown to depend heavily on the neighboring DNA bases and whether the base is at the end of a molecule or at the center of a molecule.

\subsection{Generative Process} We have codified the statistical structure of these errors in the following generative process (we have used the mean-precision parameterized of the beta distribution):

\begin{enumerate}
	\item Choose a baseline error rate parameter $\mu_0$ and a inter-location precision parameter $M_0$.
	\item For each location $j$:
	\begin{enumerate}
		\item Sample a location specific error rate parameter $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$
		\item Choose an inter-replicate precision parameter $M_j$
		\item For each replicate $i$:
		\begin{enumerate}
			\item Sample an error rate $\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$ for location $j$ in sequencing replicate $i$
			\item Sample the non-reference read count $r_{ji} \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$
		\end{enumerate}
	\end{enumerate}
\end{enumerate}

This model has three levels of sampling. First the baseline error rate across all locations and all replicates is chosen once and for all for the entire data set. Then, based on that baseline error rate, location-specific error rates are sampled. Based on the location specific error rate, replicate-specific error rates are selected. Finally, the observable data, the count of the non-reference reads is sampled.

@@

% Describe graphical model framework here

Given this model structure, the joint distribution over the latent and observed variables for data at location $j$ in replicate $i$ given the parameters is

@@

Finally, assuming the conditional independence structure in the graphical model in Figure~\ref{fig:graphical_model}, the likelihood of the data set is

@@

Instead of attempting to directly compute the posterior distribution, we sample it using a Metropolis-Hastings within Gibbs scheme as described in Section~\ref{sec:algorithm}.

@@

The primary object of inference in this model the joint posterior distribution function over the latent variables,
\begin{equation}
	p(\mu, \theta | r, n; \phi)  = \frac{ p(\mu, \theta, r | n; \phi) } {p ( r | n; \phi)},
\end{equation}
where the parameters are $\phi \triangleq \{\mu_0, M_0, M\}$.

We use a Metropolis-within-Gibbs sampling approach shown in Algorithm~\ref{alg:metro_gibbs} to approximate the posterior distribution. The model parameters are initialized using method-of-moments. Given those parameter estimates we first sample from the marginal posterior distribution for $\mu_j$ given it's Markov blanket using Metropolis Hasting. Then, we sample from the marginal posterior distribution for $\theta_{ji}$ given it's Markov blanket.

@@

\State Initialize $\theta$, $\mu$, $M$, $\mu_0$, $M_0$
\Repeat
\For {each position j} \Comment{Sample $\mu_j$}
  \State Draw T samples from $p \left( \mu_j |\theta_{ij},\mu_0,M_0\right)$ using M--H
  \State Set $\mu_j$ to the sample median for the T samples

@@

\EndFor
\Until {convergence and sample size sufficient}
\end{algorithmic}
\end{algorithm}

Once we have obtained an estimate of the posterior distribution for the case and control data, we use a two-stage test for variant at each position. First, we test whether the error rate in the case sample is significantly greater than the error rate in the control data for each position. Then, we test the hypothesis that the error rate is high is due to excess reads with a particular non-reference base. We accomplish the first test with a Bayesian posterior density test and the second with a $\chi^2$ goodness-of-fit test to a uniform distribution. In the remainder of this section, we describe the initialization, MCMC sampling approach and hypothesis testing procedure.

%%%%%%%%%%%%%
% Initialization
%%%%%%%%%%%%%
\subsection{Initialization}
The initial values for the model parameters and latent variables is obtained by a method-of-moments(MoM) procedure. MoM works by setting the population moment equal to the sample moment. A system of equations is formed such that the number of moment equations is equal to the number of unknown parameters and the equations are solved simultaneously to give the parameter estimates. We simply start with the data matrices $r$ and $n$ and work up the hierarchy of the graphical model solving for the parameters of each conditional distribution in turn.

We present the initial parameter estimates here and provide the derivations in Appendix~\ref{sec:appendix_mom}. The MoM estimate for replicate-level parameters are $\tilde{\theta}_{ji} = \frac{r_{ji}} {n_{ji}}$. The estimates for the position-level parameters are $\tilde{\mu}_j = \frac{1}{N} \sum_{i=1}^N \theta_{ji}$ and $\tilde{M_j} = \frac{ \tilde{\mu}_j (1 - \tilde{\mu}_j ) } { \frac{1}{N} \sum_{i=1}^N \theta_{ji}^2 } -1$. The estimates for the genome-level parameters are $\tilde{\mu}_0 = \frac{1}{J} \sum_{j=1}^J \mu_j$ and $\tilde{M}_0 = \frac{ \tilde{\mu}_0 (1 - \tilde{\mu}_0 ) } {\frac{1}{J} \sum_{j=1}^J \mu_j^2 } -1$.

The method-of-moments is known to have the potential to give estimates that are outside the true parameter bounds. For example, the MoM estimate $\tilde{M}_j$ may be negative while $M_j$ must be non-negative. We have not encounter such pathological behavior of the estimates in this application.

\subsection{Sampling from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$}\

@@

\begin{equation}
	p(\theta_{ji} | r_{ji}, n_{ji}, \mu_j, M_j) \thicksim \text{Beta}\left( \frac{r_{ji} + M_j \mu_j}{n_{ji} + M_j} , n_{ji} + M_j\right),
\end{equation}
where we have used the mean-precision parameterization of the beta distribution.

\subsection{Sampling from $p \left( \mu_j |\theta_{ji},\mu_0,M_0\right)$}

The posterior distribution for $\mu_j$ is
\begin{equation}
	p \left( \mu_j |\theta_{ij},\mu_0,M_0\right) = 
\end{equation}
\subsection{Bayesian hypothesis test for high error rate}

To detect a mutation in a specific position, a threshold $\tau$ is set for the posterior difference distribution testing.
\begin{align}
 H_1: \mu_{case}-\mu_{control}\geq\tau \\
 H_2: \mu_{case}-\mu_{control}<\tau
\end{align}

We can calculate the posterior probability of the hypothesis $H_1$ being true by integrating the posterior density over the correct region:

\begin{equation}
 p(H1:\mu_{case}-\mu_{control}\geq\tau)=\int_\tau^\infty p(\mu_{diff} |r,n; \mu_0,M_0)d\mu_{diff}
\end{equation}

We accept the hypothesis $H_1$ if this posterior probability is no lower than 95\%, which means we are 95\% sure that there is a mutation in this position.

ROC curve is plotted to evaluate the performance of the detection algorithm with different threshold for different mutation rate.

\subsection{$\chi^2$ test for non-uniform base distribution}

We assume that at one position in one experimental replicate, the non-reference reads follows multinomial distribution of $k = 3$ categories, with probability vector $p = (p_1, p_2, p_3)$ where $p_i \geq 0$ for $i =1, 2, 3$ and $\sum_{i=1}^3 p_i = 1$.

The non-reference reads can be because of either mutation or various errors including sampling error, inter-replicated error, $etc$. In theory, the non-reference read counts caused by non-oriented errors should be uniformly distributed among three non-reference categories, namely $p_1=p_2=p_3=1/3$. However, the distribution of non-reference read counts among three non-reference categories caused by mutation theoretically is biased.

The null hypothesis can be denoted as
\begin{align}
 H_0: p=p_0
\end{align}
where
\begin{align} 
 p_0=(p_1,p_2,p_3), p_1=p_2=p_3=1/3
\end{align}

The null hypothesis is tested using power-divergence family of statistics proposed by Cressie and Read (1984).By definition, the power-divergence family of statistcs parameterized with $\lambda$ is 
\begin{equation}
 2nI^\lambda = \frac{2}{\lambda(\lambda+1)}\sum_{i=1}^k r_i \left[\left(\frac{r_i}{E_i}\right)^\lambda-1\right];\lambda \in R,
\end{equation}
where $\{r_i; i=1,...,k\}$ are the observed frequencies and $\{E_i; i=1,...,k\}$ are the  expected frequencies.

Person's $\chi^2 (\lambda = 1)$ statistics, the log likelih ratio statistics $(\lambda = 0)$, the Freeman-Tukey statistics $(\lambda = -1/2)$ the Neyman modified $X^2 (\lambda = -2)$ are special cases of the family. Cressie and Read (1984) also proposed that when the $\lambda =2/3$, the new power-divergence statistics outcompete the traditional power-divergence statistics.

Particulaly in our case, Let $(r_1,r_2,r_3)$ denotes the non-reference base frequencies and $\sum_{i=1}^3 r_i = r$. The expected frquencies are $(rp_1, rp_2, rp_3)$. Then the power-divergence statistics can be written as

\begin{equation}
 2nI^\lambda = \frac{2}{\lambda(\lambda+1)}\sum_{i=1}^3 r_i \left[\left(\frac{r_i}{rp_i}\right)^\lambda-1\right];\lambda \in R
\end{equation}

We use $\lambda =2/3$ for the goodness of fit testing.

For each position we have six power-divergence statistics from 6 experimental replicates. We use Fisher's method to combines the six statistics into a single one for inference. Bonferroni correction is applied to maintaining the familywise error rate (FWER). The significance level is chosen as 0.05.

\subsection{Posterior distribution difference sampling}\

Using MCMC sampling, we can obtain the posterior distribution $p \left( \mu |\theta,\mu_0,M_0\right)$ for reference sample and testing sample. In order to compare the difference of position error rate $\mu$ between reference sample and testing sample , a random sampling of posterior distribution is done

\begin{enumerate}
 \item Randomly  draw one Gibbs sample from $p \left( \mu |r, n, M, \mu_0, M_0\right)$ from reference samples
 \item Randomly draw one Gibbs sample from $p \left( \mu |r, n, M, \mu_0,M_0\right) $ from testing samples
 \item Subtract the former from the later to get one sample for the posterior difference distribution 
\end{enumerate}
Repeat the above three steps until sample size is sufficient.

With the posterior difference distribution of position error rate $\mu$ across replicates obtained, Bayesian inference can be done to summarize the difference.

@@

\subsection{Overall of the process}
Figure 1

\subsection{Performance with read depth}\
Figure 2 ROC plot showing performance with different read depth

\begin{figure}[h]
\begin{center}
\includegraphics[width=160mm]{pdf_figs/ROC_without_chi2.pdf}
\caption{ROC curve varying read depth showing detection performance of model with Bayesian Hypothesis Test.}
\label{fig:ROC_without_chi2}
\end{center}
\end{figure}

\begin{figure}[h]
\begin{center}
\includegraphics[width=160mm]{pdf_figs/MuBarPlot.pdf}
\caption{ROC curve varying read depth showing detection performance of model with Bayesian Hypothesis Test.}
\label{fig:ROC_without_chi2}
\end{center}
\end{figure}

We tested the mutation detection power of our graphical model on synthetic dataset using a two-stage test scheme with dilution rates $0.1\%$, $0.3\%$, $1.0\%$, $10.0\%$ and read depth varying from $10^2$ to $10^5$.Different read depth dataset was achieve by thining dataset with read depth up to $10^6$ using Picard. $\chi^2$ test, the second stage test can be optionally included based on average read depth.

Excluding $\chi^2$ test, the ROC curves of our graphical model using Bayesian Hypothesis Testing with different threshold $\tau$ is shown in Figure??. The model had high detection power when the read depth was at $10^5$ level. At dilution rate at $0.1\%$, the model achieved performance with sensitivity at 99.74\% and specificity at 92.86\%. We missed one mutation at position 205 and had one false positive at position 8. The threshold was chosen at 0.022\% which achieved the optimal sensitivity and specificity in the ROC curve. As expected, the detection power of the graphical model decreased when the read depth was reduced.

\begin{figure}[h] %needs to go to supplementary document.
\begin{center}
\includegraphics[width=160mm]{pdf_figs/ROC_subplots_with_chi2.pdf}
\caption{ROC curve varying read depth showing method performance with Bayesian Hypothesis Testing and Chi-Square Testing}
\label{fig:R}
\end{center}
\end{figure}

Figure ?? shows the ROC curves when both Bayesian Hypothesis Testing and $\chi^2$ testing are carried out for variant calling. Comparing with Figure ??, it can be seen that when read depth is low (from $10$ to $10^3$), the false positive rate is severely limited to zero, which means we are very confident in our variant callings. However, the low false positve rate can be achieved at a sacrifice of low true postive rate especially when the dilution rate and read depth are low. The $\chi^2$ test is a stringent test when the read depth is low.

However, if the read depth is high enough (up to $10^4$ or higher), the $\chi^2$ test can limit the false positive rate in a slight degree but no longer at a sacrifice of True Positive Rate. Actually the performance is not much influenced by $\chi_2$ testing with high read depth.

@@

Figure 3 ROC comparison with other methods: varscan/samtools/rvd

\subsection{Empirical Results on Multiple Sclerosis Data}\
We tested our method on sequence data from clinical Multiple Sclerosis samples. We expect to see that

Table: Table of Variants

>>>>

\usepackage{verbatim}
\usepackage{amssymb}
\usepackage{amsmath}

\usepackage{enumitem}

\usepackage{setspace}
\doublespacing
%\usepackage{epstopdf}
%\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

@@

% DIscuss low-purity samples and somatic mutations

Next-generation sequencing (NGS) technology has enabled the genomic interrogation for a fraction of the cost of traditional assays~\cite{Koboldt2013Cell}. Protocol and platform engineering improvements have reduced the cost of the assay to the point where it is now possible to generate $1\times10^9$ bases of sequence data in under 40 hours for less than \$1000 \cite{}. As a result NGS is increasingly being used as a general platform for research assays for methylation state \cite{}, DNA mutations \cite{}, copy number variation \cite{}, promoter occupancy \cite{} and others \cite{}. Recently, NGS has been used as a clinical diagnostic tool for testing for mutations in non-invasive fetal DNA \cite{}, infectious disease samples \cite{}, cancerous tumors \cite{}, and human microbial communities \cite{}.

Clinical samples are often genetically heterogeneous. Non-invasive fetal DNA testing uses a sample of the maternal blood and filters for the minor fraction that is cell-free DNA from the fetus \cite{}. Infectious diseases such as HIV and influenza mutate rapidly and a given sample from an individual host may contain many sub-populations \cite{}. DNA sequencing of individual regions of a solid tumor has revealed genetic heterogeneous within an individual sample.

However, the primary statistical tools for calling variants from NGS data are optimized for homogeneous samples. Most analysis pipelines make use of preprocessing or postprocessing or both to eliminate error prone reads and false positives. Samtools/bcftools use a relatively straightforward XXX to call genotypes \cite{}. GATK uses a naive Bayesian decision rule to call a variant as one of two possible alleles \cite{McKenna2010}. VarScan2-mpileup uses information from multiple replicates to call homogeneous and heterogeneous genotypes.

Recently, researchers have developed algorithms to call low-frequency or rare variants in heterogeneous samples. MuTect is able to handle low-purity samples by computing a log-odds score. If the likelihood of the data assuming the locus is mutated exceeds the likelihood of the data assuming the site is not mutated by a threshold then a variant is called. There is an additional log-odds check on the normal sample to filter false positives~\cite{Cibulskis2013}. Strelka models the joint distribution for a tumor and normal sample as a mixture model using continuous values allele frequencies \cite{}. VarScan2 has a subcommand to call somatic mutations \cite{}. Recently VarScan2 was shown to outperform many other algorithms for calling low-frequency (5\%) variants \cite{}.

The remainder of this paper presents a novel algorithm for calling rare variants in heterogeneous samples using NGS data. We present the hierarchical Bayesian model for the method in Section X and the Bayesian hypothesis test to detect mutations in Section X. In Section X we present sensitivity and specificity results of our method on pure DNA samples mixed at defined fractions. We compare our algorithm to the most accurate methods available to date in Section X. Finally, in section X we present results on detecting mutations associated with multiple sclerosis from data obtained from pooling 100 affected patients. This strategy effectively reduces the cost of the assay by $n$-fold where $n$ is the pool size.

\section{Materials and Methods}

RVD uses a two-stage approach for detecting for rare variants. First, it estimates the parameters of two hierarchical Bayesian models; one using data from the sample of interest (case) and one using data from a known reference sample (control). Then, it tests for a significant difference between model parameters in the case and control samples and returns called variant positions.

\subsection{Model Structure}

For a given sample, the observed data consists of two matrices $r \in \RR^{J \times N}$ and $n \in \RR^{J \times N}$, where $r_{ji}$ is the number of reads with a non-reference base at location $j$ in experimental replicate $i$ and $n_{ji}$ is the total number of reads at location $j$ in replicate $i$. We model the data generation process as a hierarchical Bayesian model; independent models are fit to the case and control data. The generative process is as follows:

\begin{enumerate}[noitemsep]
	\item For each location $j$: 
	\begin{enumerate}
		\item Draw an error rate $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$
		\item For each replicate $i$:
		\begin{enumerate}
			\item Draw $\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$
			\item Draw $r_{ji} | n_{ji} \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$
		\end{enumerate}
	\end{enumerate}
\end{enumerate}

This process involves several hyperparameters: $\mu_0$, a global error rate; $M_0$, a global precision; $M_j$, a local precision. The global error rate, $\mu_0$, estimates the expected error rate across all locations. The global precision, $M_0$, estimates the variation in the error rate  across locations. The local precision, $M_j$, estimates the variation in the error rate across replicates at location $j$.

RVD2 has three levels of sampling. First, a global error rate and global precision are chosen once for the entire data set. Then, at each location, a local precision is chosen and a local error rate is sampled from a beta distribution. Finally, the error rate for replicate $i$ at location $j$ is drawn from a beta distribution and the number of non-reference reads is drawn from a binomial.

RVD2 hierarchically partitions sources of variation in the data. $r_{ji} | n_{ji} \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$ models the variation due to sampling the pool of DNA molecules on the sequencer. $\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$ models the variation due to experimental reproducibility. The variation in error rate due to sequence context is modeled by $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$. Importantly, increasing the read depth $n_{ji}$ only reduces the sampling error, but does nothing to reduce experimental variation or variation due to sequence context.

@@

Figure~\ref{fig:graphical_model} shows a graphical representation of the RVD2 statistical model. In this graphical model framework a shaded node represents an observed random variable, an unshaded node represents an unobserved or latent random variable and a directed edge represents a functional dependency between the two connected nodes~\cite{}. A rounded box or ``plate" represents replication of the nodes within the plate. This graphical model framework connects graph theory and probability theory in a way that facilitates algorithmic methods for statistical inference.

The joint distribution over the latent and observed variables for data at location $j$ in replicate $i$ given the parameters is

@@

Finally, the complete likelihood is

@@

@@

The primary object of inference in this model is the joint posterior distribution function over the latent variables,
\begin{equation}
	p(\mu, \theta | r, n; \phi)  = \frac{ p(\mu, \theta, r | n; \phi) } {p ( r | n; \phi)},
\end{equation}
where the parameters are $\phi \triangleq \{\mu_0, M_0, M\}$.

Exact inference in this model is intractable due to the coupling between the local and global beta distributions. [XXX complete this proof]

Instead, we have developed a Metropolis-within-Gibbs  approximate inference algorithm shown in Algorithm~\ref{alg:metro_gibbs}. First, the hyperparameters are initialized using method-of-moments. Given those hyperparameter estimates, we sample from the marginal posterior distribution for $\mu_j$ given its Markov blanket using a Metropolis-Hasting rejection sampling rule. Finally, we sample from the marginal posterior distribution for $\theta_{ji}$ given its Markov blanket. Samples from $\theta_{ji}$ can be drawn from the posterior distribution directly  because the prior and likelihood form a conjugate pair. This sampling procedure is repeated until the chain converges to a stationary distribution then we draw samples from the posterior distribution over latent variables.

@@

\State Initialize $\theta$, $\mu$, $M$, $\mu_0$, $M_0$
\Repeat
\For {each location j} \Comment{Sample $\mu_j$}
  \State Draw T samples from $p \left( \mu_j |\theta_{ij},\mu_0,M_0\right)$ using M--H
  \State Set $\mu_j$ to the sample median for the T samples

@@

\EndFor
\Until {sample size sufficient}
\end{algorithmic}
\end{algorithm}

%%%%%%%%%%%%%
% Initialization
%%%%%%%%%%%%%
\subsection{Initialization}
The initial values for the model parameters and latent variables is obtained by a method-of-moments (MoM) procedure. MoM works by setting the population moment equal to the sample moment. A system of equations is formed such that the number of moment equations is equal to the number of unknown parameters and the equations are solved simultaneously to give the parameter estimates. We simply start with the data matrices $r$ and $n$ and work up the hierarchy of the graphical model solving for the parameters of each conditional distribution in turn.

We present the initial parameter estimates here and provide the derivations in Appendix~\ref{sec:appendix_mom}. The MoM estimate for replicate-level parameters are $\tilde{\theta}_{ji} = \frac{r_{ji}} {n_{ji}}$. The estimates for the position-level parameters are $\tilde{\mu}_j = \frac{1}{N} \sum_{i=1}^N \theta_{ji}$ and $\tilde{M_j} = \frac{ \tilde{\mu}_j (1 - \tilde{\mu}_j ) } { \frac{1}{N} \sum_{i=1}^N \theta_{ji}^2 } -1$. The estimates for the genome-level parameters are $\tilde{\mu}_0 = \frac{1}{J} \sum_{j=1}^J \mu_j$ and $\tilde{M}_0 = \frac{ \tilde{\mu}_0 (1 - \tilde{\mu}_0 ) } {\frac{1}{J} \sum_{j=1}^J \mu_j^2 } -1$. While MoM is known to have the potential to return estimates outside the true parameter bounds, we have not encounter such pathological behavior in this application.

%%%%%%%%%%%%%%%%%%
% Sampling theta
%%%%%%%%%%%%%%%%%%
\subsection{Sampling from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$}

@@

\begin{equation}
	p(\theta_{ji} | r_{ji}, n_{ji}, \mu_j, M_j) \thicksim \text{Beta}\left( \frac{r_{ji} + M_j \mu_j}{n_{ji} + M_j} , n_{ji} + M_j\right).
\end{equation}

%%%%%%%%%%%%%%%%%%
% Sampling mu
%%%%%%%%%%%%%%%%%%
\subsection{Sampling from $p \left( \mu_j |\theta_{ji},\mu_0,M_0\right)$}
The posterior distribution over $\mu_j$ given its Markov blanket is 
\begin{equation}
	p( \mu_j | \theta_{ji}, M_j, \mu_0, M_0 ) \propto p(\mu_j | \mu_0, M_0) p(\theta_{ji} | \mu_j, M_j).
\end{equation}

Since the prior, $p(\mu_j | \mu_0, M_0)$, is not conjugate to the likelihood, $p(\theta_{ji} | \mu_j, M_j)$, we cannot write an analytical form for the posterior distribution. Instead, we sample from the posterior distribution using the Metropolis-Hastings algorithm.

A candidate sample is generated from the symmetric proposal distribution $Q(\mu_j^* | \mu_j^{(p)}) \thicksim $. The acceptance probability is then
\begin{equation}
	a = \frac{ p(\mu_j^* | \mu_0, M_0) p(\theta^{(p+1)}_{ji} | \mu_j^*, M_j) } {p(\mu_j^{(p)} | \mu_0, M_0) p(\theta^{(p+1)}_{ji} | \mu_j^{(p)}, M_j)}
\end{equation}

Since, each position $j$ is exchangeable given the global hyperparameters $\mu_0$ and $M_0$ this algorithm can be distributed across $J$ processors. We have found that the algorithm performance improves when we take the median of ten or more M-H samples as a single Gibbs step for each position. This computational effort only minimally increases run time when run on a parallel computing platform. In the experiments, we use a XX sample burn-in to allow the Markov chain to approach its stationary distribution and we thin the chain by a factor of 2 to reduce autocorrelation among samples.

%%%%%%%%%%%%%%%%%%
% Posterior Density Test
%%%%%%%%%%%%%%%%%%
\subsection{Hypothesis Test}
Metropolis-within-Gibbs provides samples from the posterior distribution of $\mu_j$ for both the control and case data. For notational simplicity, we define the random variables associated with these two distributions $\tilde{\mu}_j^{\text{case}}$ and $\tilde{\mu}_j^{\text{control}}$.

A variant is called if $\tilde{\mu}_j^{\text{case}} > \tilde{\mu}_j^{\text{control}}$ with high confidence,
\begin{equation}
	\Pr( \tilde{\mu}_j^{\text{case}} - \tilde{\mu}_j^{\text{control}} \geq \tau ) > 1-\alpha,
\end{equation}
where $\tau$ is a detection threshold and $1-\alpha$ is a confidence level. We draw a sample from the posterior distribution $\tilde{\mu}_j^{\Delta} \triangleq \tilde{\mu}_j^{\text{case}} - \tilde{\mu}_j^{\text{control}}$ by randomly sampling with replacement from the samples of $\tilde{\mu}_j^{\text{case}}$ and $\tilde{\mu}_j^{\text{control}}$.

\section{Comparison with other variant calling algorithms}

GATK uses considerable preprocessing and a simple naive Bayes decision rule to call a variant. The genotype prior is fixed and constant across all loci and the likelihood of an allele at a locus is a function of the phred score \cite{McKenna2010}. Samtools

%%%%%%%%%%%%%%%%%%

% Chi^2 Test

%%%%%%%%%%%%%%%%%%
%\subsection{$\chi^2$ test for non-uniform base distribution}
%
%We assume the counts for the non-reference bases at poisiuton $j$ in replicate $i$ follow a multinomial distribution with parameter  $p = (p_1, p_2, p_3)$ where $p_k \geq 0$ for $k =1, 2, 3$ and $\sum_{k=1}^3 p_k = 1$.
%
%The non-reference reads may be due to a true mutation or due to a random sequencing error. We assume, non-reference read counts caused by a random mechanism results in a uniform distributed among three non-reference categories, giving $p_1=p_2=p_3=1/3$. In contrast, the distribution of non-reference read counts among three non-reference categories caused by mutation would be biased to one base. 
%
%We use a $\chi^2$ goodness-of-fit test to distinguish these two possible scenarios. The null hypothesis is $H_0: p = (p_1, p_2, p_3)$ where $p_1=p_2=p_3=1/3$. Cressie and Read (1984) identified a power-divergence family of statistics, indexed by $\lambda$, that includes as special cases Pearson's $\chi^2 (\lambda = 1)$ statistics, the log likelihood ratio statistics $(\lambda = 0)$, the Freeman-Tukey statistics $(\lambda = -1/2)$ the Neyman modified $X^2 (\lambda = -2)$. The test statistic is
%
%\begin{equation}
% 2nI^\lambda = \frac{2}{\lambda(\lambda+1)}\sum_{k=1}^3 r_{ji}^{(k)} \left[\left(\frac{r_{ji}^{(k)}}{E_{ji}^{(k)}}\right)^\lambda-1\right];\lambda \in R,
%\end{equation}
%
%where $r_{ji}^{(k)}$ is the observed frequency for non-reference base $k$ at position $j$ in replicate $i$ and $E_{ji}^{(k)}$ is the corresponding expected frequency under the null hypothesis. Cressie and Read (1984) proved that when $\lambda = 2/3$ this statistic is XXX.
%
%For each position, there are $N$ power-divergence statistics for each experimental replicate. We use Fisher's combined probability test to combine the p-values from each replicates into a single one for each position,
%
%\begin{equation}
%	X^2 = -2 \sum_{i=1}^N \ln(p_i).
%\end{equation}
%This gives a test statistic that follows a $\chi^2$ distribution with $2*N$ degrees of freedom when the null hypothesis is true.
%
%While Fisher's combined probability test collapses the p-values across replicates, we still have $J$ tests across positions. We use the Bejamini-Hochberg method to control the family-wise error rate (FWER) across positions.

@@

\subsection{Performance with read depth}\

We synthesized a 400bp control DNA sequence and another 400bp sequence that is identical to the control except at 14 loci with variant bases. We denote the indices of these true positive variant positions $j^*$. Pure case and control samples were mixed at defined fractions to yield minor allele frequencies (MAFs) of 0.1\%, 0.3\%, 1\%, 10\%, and 100\%. The sequencing protocol and data set are available from the original publication~\cite{}.

BAM files for the synthetic DNA data were down sampled by $10\times$, $100\times$, $1,000\times$, $10,000\times$ to simulate lower coverage data while retaining the error structure of real NGS data. The final data set contains pairs of reads for three replicates of each case and a pair of reads for the control sample giving a total of 36 BAM files for each downsampled data set.

We ran the RVD2 algorithm on each downsampled data set sand varied the threshold $\tau$ with fixed $\alpha=0.05$ to generate receiver operating characteristic (ROC) curves in Figure~\ref{fig:ROC}. Figure~\ref{ig:ROC}A shows ROC curves for a true 0.1\% MAF for a range of median coverage depths. At the lowest depth the sensitivity and specificity is no better than a random decision. We would not expect to be able to call a 1 in 1000 variant base with a coverage of only 43. The performance improves monotonically and is much better at a depth of approximately 41,000. This depth exceed the inverse MAF by a $10\times$ which we find to be generally a good coverage threshold for detecting variants. We find a similar pattern in Figures~\ref{fig:ROC}B-C. Interestingly, the algorithm performance is much better than a random decision rule to detect a 10\% MAF with only a depth of 25.

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=80mm]{pdf_figs/ROC_without_chi2.pdf}
\caption{ROC curve varying read depth showing detection performance of model with Bayesian Hypothesis Test.}
\label{fig:ROC}
\end{center}
\end{figure}

Figure~\ref{fig:MAF} shows the posterior mean and 95\% credible intervals for $\mu_j$ for called variant positions with $\bar{n} = 41,573$. The only false positive is at position 8, though posterior distributions for the case and control clearly show a difference at that position. The posterior mean estimates are shrunken towards the global error rate parameter $\mu_0 = 0.0023$ for the case and control due to the empirical Bayes procedure.

\begin{figure}[h]
\begin{center}
\includegraphics[width=80mm]{pdf_figs/MuBarPlot.pdf}
\caption{Estimated minor allele fraction for called variants in 0.1\% dilution.}
\label{fig:MAF}
\end{center}
\end{figure}

@@

Figure 3 ROC comparison with other methods: varscan2/samtools/rvd/strelka/virmid/muTect

\subsection{Pooled Multiple-Sclerosis Samples}\

We tested our algorithm on data from pooled samples from patients with multiple-sclerosis (MS). The MS patients are ``high-burden"; meaning that they are known to carry one or more of several genetic risk alleles. Our aim with studying these patients is to identify rare risk alleles that may not yet be known to be associated with MS.

DNA was extracted from whole blood samples from 100 patients with multiple sclerosis and 100 unaffected, unrelated individuals. Each sample was subjected to targeted PCR amplification of IL2RA and interleukin 7 receptor (IL7R); genes with known risk alleles~\cite{HauserNEJM2007}. We targeted X eons covering XXbp in IL2RA and X exon covering XXbp in IL7R. The experiment was repeated on two independent cohorts to filter for spurious associations due to population sampling.

We took 4000 samples from the Markov chain, removed the first 10\% for burn-in and thinned the chain by a factor of 2. We used 10 Metropolis samples for each Gibbs step. We called variants with $\tau = 0.005$ and $\alpha = 0.05$ because we expect that the minimum detectable genotype is a heterozygous mutation in 1 out of 100 samples.

We identified by consensus of the two independent cohorts two variants in IL7R and one variant in IL2RA. All position information is referenced to human genome reference hg19. One variant in IL7R is rs1494558 at chr5:35861068 T$>$C and has a MAF of 76\% and 71\% in the two cohorts. This mutation causes a missense variant Ile66Thr and has been shown to be associated with abnormal T-cell response and IgA neuropathy~\cite{HahnCho2011}. The second IL7R variant is at chr5:35875593 A$>$T and has a MAF of 60\% and 55\% in the two cohorts. This variant has not yet been shown to be associated with any known disease. The IL2RA variant is at chr10:6061965 A$>$T and has a MAF  of 3\% and 5\% in the two cohorts. This variant has not yet been shown to be associated with any known disease.

\begin{figure}[h]
\begin{center}
%\includegraphics[width=80mm]{pdf_figs/MuBarPlot.pdf}
\caption{Estimated minor allele fractions for consensus variants in MS patient pools.}
\label{fig:MS_MAF}
\end{center}
\end{figure}

The validity of these loci are subject to confirmation on a larger independent sample. However, we note that most GWAS studies focus on common SNPs due to the cost-effective microarray technology used for those studies. We have instead used sequencing technology which yields and unbiased coverage of the targeted regions and has the potential to uncover rare variants that are not targeted by microarray technology. Indeed, the common SNP that was found is associated with an autoimmune disorder and the other two variants are not at dbSNP loci.
